On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-23 paper/paper3d.tex
minor edits

Despite this, the Toomre parameter and its associated
dispersion relation are still widely applied to viscous, cooling
accretion disk models \citep[e.g.][]{cossins09, kimura12}, 
even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
derived from inviscid disks without cooling.

>>>>

Despite this, the Toomre parameter and its associated
dispersion relation are still widely applied to viscous, cooling
accretion disk models \citep[e.g.][]{cossins09, kimura12}, {\bf more refs?}
even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
derived from inviscid disks without cooling.

### 2016-01-23 paper/results2d.tex
minor edits

\section{Two-dimensional disks with beta cooling}\label{result_2d}
We begin in the 2D limit with beta cooling  to make connection 
with previous studies. The disk
material is assumed to be confined to the midplane so $\delta v_z=0$. 
We make the replacement  
$\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
pressure, and set $\Gamma=1$ without loss of generality. 
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

The linearized equations yield an algebraic dispersion relation $s =
s(k)$. We write this  
in terms of the dimensionless growth rate $S = s/\Omega$ and
wavenumber $K=kH = k c_{s0}/\Omega$ as
\begin{align}\label{thindisk}
  f(S,K)\equiv AD - BC = 0,   
\end{align}
where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}.

\subsection{Inviscid limit}\label{2d_inviscid}
Here we neglect viscosity by setting 
$\alpha = \alpha_b = 0$. This eliminates viscous GI and 
allows us to quantify the \emph{sole} effect of cooling on the
perturbations. An external, time-independent heat source should be
invoked to balance the imposed cooling to allow an equilibrium to be
defined ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are
assuming viscosity only provides a background heating and does not
play an active role in the perturbed state. %may or may not be true in a real
                                %disk because we are using viscosity
                                %as proxy for turbulence

Eq. \ref{thindisk} simplifies to  
\begin{align}\label{inviscid}
%  &\beta S^3 + S^2 - \beta \left\{ \frac{2|K|}{\qtwo} - \left[2(2-q) +
%    \gamma K^2\right]\right\}S \notag\\ &- \left[\frac{2|K|}{\qtwo} - 2(2-q)\right]
%  = 0. 
  S^2 = \frac{2|K|}{Q} - 2(2-q) - \left(\frac{\theta + \beta \gamma
    S}{1+\beta S}\right)K^2, 
\end{align}
similar to the classic Lin-Shu dispersion relation, which may be
obtained by taking the limit $|\beta S|\to\infty$. 
The first term on
the right-hand-side represents destabilization by self-gravity; 
the second and third terms represent stabilization by rotation and
pressure, respectively. The imposed cooling only affects the
pressure response.

>>>>

\section{Two-dimensional disks with beta cooling}\label{result_2d}
We begin in the 2D limit with beta cooling  to make connection 
with previous studies. The disk
material is assumed to be confined to the midplane so $\delta v_z=0$. 
We make the replacement  
$\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
pressure, and set $\Gamma=1$. % without loss of generality. 
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

The linearized equations yield an algebraic dispersion relation $s =
s(k)$. We write this  
in terms of the dimensionless growth rate $S = s/\Omega$ and
wavenumber $K=kH = k c_{s0}/\Omega$ as
\begin{align}\label{thindisk}
  f(S,K)\equiv AD - BC = 0,   
\end{align}
where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}. 
We use this generalized dispersion relation to investigate GI driven
solely by cooling in \S\ref{2d_inviscid}; and GI driven by viscosity
in \S\ref{2dvisc}.

\subsection{Inviscid limit}\label{2d_inviscid}
We first simplify the problem by setting $\alpha = \alpha_b = 0$. This
eliminates viscous effects and allows us to quantify the 
\emph{sole} effect of cooling on the 
perturbations. An external, time-independent heat source should be
invoked to balance the imposed cooling to allow an equilibrium to be
defined ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are
assuming viscosity only provides a background heating and does not
play an active role in the perturbed state. %may or may not be true in a real
                                %disk because we are using viscosity
                                %as proxy for turbulence
                                %maybe turb heating actually behaves
                                %like background heat source only 
Eq. \ref{thindisk} becomes 
\begin{align}\label{inviscid}
%  &\beta S^3 + S^2 - \beta \left\{ \frac{2|K|}{\qtwo} - \left[2(2-q) +
%    \gamma K^2\right]\right\}S \notag\\ &- \left[\frac{2|K|}{\qtwo} - 2(2-q)\right]
%  = 0. 
  S^2 = \frac{2|K|}{Q} - 2(2-q) - \left(\frac{\theta + \beta \gamma
    S}{1+\beta S}\right)K^2, 
\end{align}
similar to the classic Lin-Shu dispersion relation
(i.e. Eq. \ref{classic_dr}), which may be obtained by taking the limit
$|\beta S|\to\infty$. %with gamma=1
The first term on the right-hand-side represents destabilization by self-gravity; 
the second and third terms represent stabilization by rotation and
pressure, respectively. The imposed cooling only affects the
pressure response.
